# Survival Analysis Review

## PDF, CDF, and Survival

$$
f(x) = \frac{1}{\lambda} e^{\frac{x}{\lambda}}; \quad X \sim Exp(\lambda = 1.5)
$$

```{r}
#| fig-align: "center"
ggplot() +
  stat_function(fun = \(x) dexp(x, rate = 1 / 1.5), 
                xlim = c(0, 10.2), linewidth = 1.5, fill = "#3f1f69", geom = "area", alpha = 0.5) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1.05)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 10.2), n.breaks = 8) +
  theme_void() +
  theme(axis.text = element_text(size = 10)) +
  labs(y = NULL)
```


## PDF, CDF, and Survival

$$
F(x) = \int_{t = 0}^{x} f(t); \quad f(x) = \frac{1}{\lambda} e^{\frac{x}{\lambda}}; \quad X \sim Exp(\lambda = 1.5)
$$

```{r}
#| fig-align: "center"
ggplot() +
  stat_function(fun = \(x) pexp(x, rate = 1 / 1.5),
                xlim = c(0, 10.2), linewidth = 1.5, fill = "#3f1f69", geom = "area", alpha = 0.5) +
   geom_hline(yintercept = 1, linetype = "dashed", color = "#aeaeb2", linewidth = 1.5) +
 scale_y_continuous(expand = c(0, 0), limits = c(0, 1.05)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 10.2), n.breaks = 8) +
  theme_void() +
  theme(axis.text = element_text(size = 10)) +
  labs(y = NULL)
```


## PDF, CDF, and Survival

$$
S(t) = 1 - F(t); \quad F(x) = \int_{t = 0}^{x} f(t); \quad f(x) = \frac{1}{\lambda} e^{\frac{x}{\lambda}}; \quad X \sim Exp(\lambda = 1.5)
$$

```{r}
#| fig-align: "center"
ggplot() +
  stat_function(fun = \(x) 1 - pexp(x, rate = 1 / 1.5),
                xlim = c(0, 10.2), linewidth = 1.5, fill = "#3f1f69", geom = "area", alpha = 0.5) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "#aeaeb2", linewidth = 1.5) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1.05)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 10.2), n.breaks = 8) +
  theme_void() +
  theme(axis.text = element_text(size = 10)) +
  labs(y = NULL)
```

## How to estimate this distribution?

$$
f(x) = \frac{1}{\lambda} e^{\frac{x}{\lambda}}; \quad X \sim Exp(\lambda = 1.5)
$$

```{r}
#| fig-align: "center"
ggplot() +
  stat_function(fun = \(x) dexp(x, rate = 1 / 1.5), 
                xlim = c(0, 10.2), linewidth = 1.5, fill = "#3f1f69", geom = "area", alpha = 0.5) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1.05)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 10.2), n.breaks = 8) +
  theme_void() +
  theme(axis.text = element_text(size = 10)) +
  labs(y = NULL)
```

## What if I can only follow for 2 years?

$$
f(x) = \frac{1}{\lambda} e^{\frac{x}{\lambda}}; \quad X \sim Exp(\lambda = 1.5)
$$

```{r}
#| fig-align: "center"
ggplot() +
  stat_function(fun = \(x) dexp(x, rate = 1 / 1.5), 
                xlim = c(0, 10.2), linewidth = 1.5, fill = "#3f1f69", geom = "area", alpha = 0.5) +
  geom_vline(xintercept = 2, linetype = "dashed", color = "#aeaeb2", linewidth = 1.5) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1.05)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 10.2), n.breaks = 8) +
  theme_void() +
  theme(axis.text = element_text(size = 10)) +
  labs(y = NULL)
```

## What if I can only follow for 2 years?

```{r}
tibble(
  id = factor(1:6),
  time = 1:6,
  dead = ifelse(time %% 2 == 0, "Died", "Censored")
) |> 
  ggplot() +
  geom_segment(aes(y = id, yend = id, x = 0, xend = time), 
               linewidth = 2.5, alpha = 0.5) +
  geom_point(aes(x = time, y = id, shape = dead), size = 3.5) +
  scale_shape_manual(values = c(1, 4)) +
  scale_x_continuous(n.breaks = 7, expand = c(0, 0.04)) +
  theme_classic() +
  labs(
    y = "Patient ID",
    x = "Time",
    shape = NULL
  )
```

## What if I can only follow for 2 years?

* A key feature of survival data is **censoring**

* A subject may be censored due to:

  * Loss to follow-up
  
  * Withdrawal from study
  
  * No event by end of fixed study period
  
* Survival analysis techniques provide a way to appropriately account for censored patients in the analysis

* The Kaplan-Meier estimate of survival probability at a given time is the product of these conditional probabilities up until that given time.
